# Chapter 3: Logistic Regression

*Performing logistic regression analysis on the student alcohol consumption data*


- We'll use the data wrangled in the first part of exercise 3: the full description of the data can be found [here](https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION)


- The data is collected on mathematics and portuguese classes: only students who answered the questionnaire on both classes have been included in the analyses.

It includes the following variables:


```{r code6, echo=T, warning = F}

# load visualization libraries

library(ggplot2)
# read the data
alc <- read.table("./data/alc_students.txt",header = T,  dec = ".")


print(colnames(alc))

```
 
 The student alcohol consumption during weekdays has been recorded in the variable Dalc, and that during weekends in Walc. The average of the two is stored in the variable alc_use. Based on this value, the stuents have been dividen in two groups: those with high alcohol use (average over 2), and the others: this is indicated in the boolean variable high_use.
 
 Next we will try to explain the high alcohol use with other variables. Now, it is not clear if he other variables cause high alcohol use, or vice versa. Still, in our model, we will explain the alcohol use using the other variables, disregarding any direction of (possible) causality. 
 
 Variables that might be related to high alcohol consumption include the student's sex, (variable sex), quality of family relationships (famrel), being in a romantic relationship (romantic), and number of school absences (absences). I would guess that, on average, males have a higher risk of high alcohol consumption. Number of school absences would possibly increase the likelihood of high alcohol use, while being in a romantic relationship and good family relationships would decrease the likelihood. Next, well explore whether these hypotheses hold true or not. 
 
 
```{r code7, echo = T}
# create the model

# initialize a plot of alcohol use
g1 <- ggplot(data = alc, aes(x = sex, fill = high_use))

# define the plot as a bar plot and draw it
g1 + geom_bar() + xlab("Sex")

```
 
 In the figure we can see that, first of all, there are a few more females compared to males. However, there are almost twice as many high alcohol users within males: so this would seem to support the initial hypothesis. We will test this further with a logistic regression model in a moment. Before that, let's look at the other variables. 
 
```{r code8, echo = T}
# create the model

# initialize a plot of alcohol use
g2 <- ggplot(data = alc, aes(x = famrel))

# define the plot as a bar plot and draw it
g2 + geom_histogram(binwidth = 1) + xlab("Quality of family relationships")

```
 
 We can see that most of the respondents have relatively good family relations. Now, let's check how these affect the alcohol consumption: 
 
```{r code9, echo = T}
# create the model


g3 <- ggplot(alc, aes(x = high_use, y = famrel))

# define the plot as a boxplot and draw it
g3 + geom_boxplot() + ylab("Quality of family relationships")


```
 
 As we can see, the box plot is not very ideal for this task: with non-high-users, the 25th percentile and median are the same value, while for the high users the same applies for the median and 75th percentile. This shows as missing halves of the boxes. Still, we can maybe see that the high users have slightly lower scores for the quality of family relationships. Next, let's look at the romantic relationship variable, and its effect on the alcohol use.

```{r code10, echo = T}
# create the model

# initialize a plot of alcohol use
g4 <- ggplot(data = alc, aes(x = romantic, fill = high_use))

# define the plot as a bar plot and draw it
g4 + geom_bar() + ylab("Being in a romantic relationship")

```

First of all, we can see that there are around twice as many people not in romantic relationship as those who are. It looks that there might be a bit smaller proportion of high alcohol users in those who are in a romantic relationship, compared to those who are not. We will see this better when we construct the model. As the last of the variables, we will look at the number of school absences. It's distribution is plotted below:

```{r code11, echo = T}
# create the model

# initialize a plot of alcohol use
g5 <- ggplot(data = alc, aes(x = absences))

# define the plot as a bar plot and draw it
g5 + geom_histogram(binwidth = 1) + xlab("Number of school absences")

```
 
So most of the people have less than ten absences. Let's see how high alcohols users compare to others in this respect:
 
```{r code12, echo = T}
# create the model


g6 <- ggplot(alc, aes(x = high_use, y = absences))

# define the plot as a boxplot and draw it
g6 + geom_boxplot() + ylab("Number of school absences")


```

Now we can see the box of the high alcohol users extending to clearly more absences. This is in line with the original hypothesis.

 